Terrain is a stronger predictor of peat depth than airborne radiometrics in Norwegian landscapes

### 2024-12-05 ms/ms.tex
ms: hyphens and abbreviations

\begin{abstract}
Peatlands are Earth's most carbon-dense terrestrial ecosystems and their carbon density varies with the depth of the peat layer.
Accurate mapping of peat depth is crucial for carbon accounting and land management, yet existing maps lack the resolution and accuracy needed for these applications.
This study evaluates whether digital soil mapping using remotely sensed data can improve existing maps of peat depth in Norway.
Specifically, we assessed the predictive value of LiDAR-derived terrain variables and airborne radiometric data at two contrasting sites.
We measured peat depth at 372 and 1878 locations and trained Random Forest models to predict depth at 10 m resolution.
The best models achieved mean absolute errors of 60 and 56 cm, explaining one-third of the variation in peat depth.
Terrain variables were better predictors than radiometric variables, with elevation and valley bottom flatness showing the strongest relationships to depth.
Radiometric data showed inconsistent value, improving predictions at one site while degrading performance at the other.
Our models improved upon the national peat depth map, but weak relationships with remotely sensed variables make peat depth hard to predict.
These findings suggest that while digital soil mapping can enhance existing peat depth maps, reliable carbon stock assessments still require strategic field sampling.
Furthermore, the inability of models to identify peat presence outside known peatlands highlights the need for integrated mapping of peat occurrence and depth.
\end{abstract}

@@

Peat soils are a terrestrial carbon stock of global importance.
They store about 30 \% of global soil carbon despite covering only 2-3 \% of Earth's land \citep{xuPEATMAPRefiningEstimates2018, friedlingsteinGlobalCarbonBudget2020, unepGlobalPeatlandsAssessment2022}.
In other words, they are extremely carbon dense -- denser than any other ecosystem per square meter \citep{temminkRecoveringWetlandBiogeomorphic2022}.
This makes peatlands crucial to climate change mitigation.
Intact peatlands sequester carbon and produce a negative temperature forcing, overall \citep{joostenRolePeatlandsClimate2016}.
When disturbed, often by conversion to another land use, they can produce large greenhouse gas emissions \citep{maGloballyRobustRelationship2022}.

@@

The most practical and widespread \emph{scorpan} factors for peat depth are relief and spatial position.
Spatial position is unique because it is always known (with varying accuracy).
However, the short range of spatial autocorrelation in peat depth limits its mapping value \citep{henglGenericFrameworkSpatial2004}.
Relief, or topography, is widely and accurately mapped in digital terrain models.
For example, most of mainland Norway has at least one elevation measurement per square meter from airborne LiDAR surveys.
Moreover, mechanisms of peat formation are linked to topography.
For example, a steep slope is unlikely to have a high water table, thereby limiting accumulation.

@@

The little we know about the depths of Norwegian peatlands comes from surveys meant to identify arable land.
After scattered surveys in the early 20th century, a comprehensive round of surveying was completed 1964-2001 as part of a wider land cover mapping in Norway \citep{bjordalMarkslagsklassifikasjonOkonomiskKartverk2007}.
Because of its agricultural and silvicultural focus, it covered only productive areas below the tree line, and peat depth was measured only in places judged to be potentially arable or afforestable \citep{ahlstromAR5Klassifikasjonssystem2019}.
Field surveyors carried a one-meter probe, so they measured peat depth as shallow (\textless{} 1 m), or deep (\textgreater{} 1 m).
These classes were assigned to whole mire polygons, so spatial resolution is on the order of hectares.

@@

At both study sites, our measurements of peat depth were made for the purpose of training a Random Forest (RF) model of peat depth, and we designed our sampling with this in mind \citep{brusSamplingDigitalSoil2019}.
Broadly, we aimed for a sample that was representative of the predictor space defined by the most important predictors of peat depth \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
A sample that preserves the properties of the multivariate distribution of predictor and outcome variables is most likely to maintain any complex, non-linear relationships that exist in the population while avoiding spurious ones \citep{brusSamplingDigitalSoil2019}.
We chose for our sampling and modelling a spatial resolution of \unit{10\,m}.
We considered this a reasonable compromise between digital terrain model (DTM) resolution (\unit{1\,m}) and small mires on the one hand, and airborne radiometric resolution (\unit{50\,m}) on the other.
The point measurements of peat depth described below were ultimately aggregated to \unit{10\,m} resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.

@@

Field work at Skrimfjella was conducted in August 2020.
We navigated to the centers of the raster cells in the depth and occurrence samples by handheld GPS, checking that positional error was below \unit{3\,m}.
For each depth sample location, we measured peat depth three times (at the vertices of a triangle with \unit{2\,m} sides) to get a more representative value for the \unit{10\,m} raster cell, and to dampen the effect of outlying measurements \citep{parryEvaluatingApproachesEstimating2014}.
We used a metal probe pushed downward until resistance indicated the base of the peat column.
Probe locations were adjusted up to \unit{20\,cm} if the base of the peat column seemed to be blocked by an obvious artifact.
A single probe measurement was right-censored because the peat column was deeper than the probe.
For each occurrence sample location, we recorded the presence or absence of peatland -- primarily by digging and examining the top \unit{20\,cm} of soil (where this was possible).
We judged whether the soil was a peat soil based on its density, texture, and color.
Occasionally, when the soil itself was difficult to judge, we made our determination also based on the presence or absence of mire vegetation.
Although peat soil is strictly defined by organic content (which we did not analyse), we believe our protocol produced reasonable determinations of presence or absence that would generally satisfy most of the varying definitions of peatland \citep{minasnyMappingMonitoringPeatland2023}.

Besides the depth and occurrence measurements described above, we also measured peat depth in three subjectively-chosen, individual mires, using ground-penetrating radar (GPR).
We used the Malå ProEx GPR system (Guideline Geo AB, Sweden) with its \unit{500\,MHz} shielded antenna mounted in a plastic sledge, and its control unit connected to a GNSS receiver.
At each of the three mires we recorded GPR traces along walking transects that covered the extent of the mire, mostly in traversing, zigzag patterns with between \unit{5\,m} and \unit{20\,m} spacing at their vertices.
Along the GPR transects we also probed peat depth at marked trace locations, to be able to calibrate the GPR wave speed velocity.
We processed the GPR data with Reflex2DQuick software (v.3.0, Sandmeier Scientific Software, Germany), applying a time-zero correction, a dewow filter, and a gain filter based on observed energy decay.
Then we picked strong reflectors in the radargrams that we interpreted as the base of the peat column.
We used picks at marked trace locations to calibrate wave speed velocity; we pooled calibration points across the three mires and fitted a linear regression of depth on one-way travel time with the intercept fixed at zero.
In total we had 46 calibration points along \unit{3.5\,km} of GPR transects.
Finally, we used the calculated wave velocity (\unit{0.0387\,m\,ns^{-1}}, \(R^2 = 0.874\)) to convert the travel times of all picks to calibrated peat depths.

@@

At Ørskogfjellet we also measured peat depth in a sample of \unit{10\,m} raster cells, selected from the part of the study area classified in the AR5 dataset as mire.
Before selecting locations, we determined a minimal sample size that would adequately capture the terrain and radiometric properties of the entire mire area.
Specifically, we aimed to identify the size at which adding locations produced diminishing decreases in divergence between sample and population distributions -- i.e.~the elbow point in a curve of similarity between sample and population \citep{maloneMethodsImproveUtility2019}.
This approach has been found to identify sample sizes that correspond with diminishing returns in predictive model performance on external evaluation data \citep{sauretteDivergenceMetricsDetermining2023}.
We defined a sequence of sample sizes (50--500) and for ten replicate samples at each size \citep[drawn by conditioned latin hypercube sampling,][]{minasnyConditionedLatinHypercube2006, roudierClhsPackageConditioned2011}, we calculated the mean Kullback-Leibler divergence between sample and population distributions \citep{maloneMethodsImproveUtility2019, sauretteDivergenceMetricsDetermining2023}.
The predictors in the divergence calculation were terrain slope and four radiometrics: potassium, thorium, uranium, and total count.
Next, we fitted a asymptotic regression of mean divergence on sample size, and identified the sample size at which the curve reached 95 \% of the fitted asymptote.
Through this procedure we found that we could adequately capture the population distribution with a sample of 160 locations.

To choose 160 locations, we performed feature space coverage sampling.
This approach has been found to produce higher accuracy in RFs than conditioned latin hypercube sampling \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
Feature space coverage sampling aims to disperse samples as uniformly as possible in multidimensional predictor space, and is implemented by choosing locations that are closest to cluster centers in a k-means clustering of the standardized predictor space \citep{brusSamplingDigitalSoil2019}.
Feature space coverage sampling works best when all dimensions are important predictors of the outcome \citep{wadouxSamplingDesignOptimization2019}, and we used the same five predictors that we used to choose sample size: terrain slope and four radiometrics.
The radiometric predictors were downscaled to \unit{10\,m} resolution with cubic B-spline resampling in QGIS \citep[v.3,][]{QGISsoftware}.
We adjusted the feature space coverage sampling to ensure that locations were accessible within time constraints, and assessed how this changed our sample from an ideal feature space coverage sample.
Adjusting for accessibility is justified because the smaller sample size that would result if accessibility were ignored can degrade model accuracy as much or more as deviations from ideal sampling designs \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
To adjust, we first restricted the sampling population to mire areas that were within an arbitrary cost distance of publicly-accessible roads.
Cost distance was calculated using GRASS's \emph{r.walk} function, with friction costs defined by AR5 land classes \citep{GRASSv8-2}.
After creating a feature space coverage sample with this restriction, we also inspected a map of the sample and substituted 16 inaccessible locations with accessible locations from the same or a nearby cluster.
Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 \% (with respect to the ideal sample), but distance in our sample was still only 46 \% of the mean distance to cluster centers -- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.

@@

We used the \emph{terra} R package to calculate from the DTMs: slope, topographic position index (difference from mean of eight neighbors), terrain ruggedness index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
These were derived at two scales to produce eight different predictors; we either calculated from \unit{1\,m} DTM resolution and then aggregated, or aggregated to \unit{10\,m} DTM resolution and then calculated the indices.
This kind of multiscale feature engineering of land surface parameters has been found to improve machine learning predictions of soil properties \citep{millerImpactMultiscalePredictor2015, dornikOptimalScalingPredictors2022, newmanAssessingSpatiallyHeterogeneous2023}.
We know that peat depth tends to vary at fine scales in Norway, which is why we chose \unit{1\,m} and \unit{10\,m} resolutions \citep{maxwellLandsurfaceParametersSpatial2022}.
We also calculated the the Multi-Resolution Valley Bottom Flatness index, which indicates the degree of valley bottom flatness at a given location via a multiscale algorithm \citep{gallantMultiresolutionIndexValley2003}.
We calculated this index in SAGA GIS \citep[v.9.3.2,][]{conradSystemAutomatedGeoscientific2015} with default parameters.

Next, we calculated the Topographic Wetness Index \citep{quinnPredictionHillslopeFlow1991}.
This index is notoriously scale-dependent and often matches real hydrological conditions best when calculated from moderate-to-coarse resolution DTMs \citep{agrenEvaluatingDigitalTerrain2014, riihimakiTopographicWetnessIndex2021}, so we calculated it from \unit{5\,m}, \unit{10\,m}, \unit{20\,m}, and \unit{50\,m} DTM resolution.
The calculations were performed with Whitebox software \citep{lindsayWhiteboxGATCase2016}, accessed through the \emph{whitebox} R package \citep[v2.4,][]{wuWhiteboxWhiteboxToolsFrontend2022}.
We filled depressions in the DTM with the algorithm in Wang \& Liu \citeyearpar{wangEfficientMethodIdentifying2006}, and used the deterministic infinity flow accumulation algorithm \citep{tarbotonNewMethodDetermination1997}.

The last terrain-based predictor we included was the depth-to-water index \citep{murphyMappingWetlandsComparison2007}.
This index approximates a location's vertical height above the surface water feature that it is likely to drain towards.
It is calculated as the minimum cumulative slope (scaled by cell size) to a surface water feature \citep[eq. 5 in][]{murphyTopographicModellingSoil2009}.
We calculated unitless slope from the \unit{1\,m} DTM using the Whitebox software.
Also using Whitebox, we defined surface water features from the DTM by filling depressions and then calculating flow accumulation to define catchment areas for each cell \citep{schonauerSpatiotemporalPredictionSoil2021, schonauerRcodeCalculatingDepthwater2021}.
This catchments area layer was then thresholded at seven different levels (\emph{flow initiation areas} from \unit{0.25\,ha} to \unit{16\,ha}) to estimate surface water features under moisture scenarios varying from wet to dry \citep{murphyModellingMappingTopographic2011, agrenEvaluatingDigitalTerrain2014, schonauerSpatiotemporalPredictionSoil2021}.
In addition, all surface water features mapped in the AR5 dataset were also transferred to the raster layer.
For each of the seven surface water layers, we derived the depth-to-water index using the \emph{Distance Accumulation} tool in ArcGIS Pro (v.3.1, ESRI, USA), which has an efficient algorithm to find the cumulative distance over a cost surface to the least-cost source.

@@

We used RF to predict peat depth at both sites.
RF is a tree-based ensemble machine learning algorithm that builds many decision trees on bootstrapped samples of the training data, randomly subsets predictors in the trees, and averages the predictions of the trees \citep{breimanRandomForests2001}.
We chose RF because it can handle complex interactions between predictors, is robust to overfitting, and generally shows higher performance in DSM applications than other algorithms \citep{beguinPredictingSoilProperties2017, nussbaumEvaluationDigitalSoil2018, lamichhaneDigitalSoilMapping2019}.
It is suited for use on relatively small training data sets and its predictions can be interrogated to learn about predictor importance \citep{khaledianSelectingAppropriateMachine2020}.
Evaluating variable importance in a maximally-predictive model aligns with the aim of this study.

RF by itself is not a spatial model, and it will only predict spatial structure in the outcome to the degree that the structure is captured by predictors.
We considered using regression kriging -- a hybrid between non-spatial and spatial techniques that would be achieved by adding to the RF predictions a geostatistically-interpolated surface of RF residuals \citep{henglGenericFrameworkSpatial2004}.
The spatial component in regression kriging often improves map accuracy compared to a non-spatial model \citep{beguinPredictingSoilProperties2017, lamichhaneDigitalSoilMapping2019, mollaMachineLearningGeostatistical2023}, but it can do so only if the spatial autocorrelation range in the non-spatial residuals is large compared to distances between samples and prediction locations \citep{henglGenericFrameworkSpatial2004, szaboMappingSoilHydraulic2019, takoutsingComparingPredictionPerformance2022}.
If the outcome varies at fine scales and the samples are clustered in small parts of the study area, a spatial component will hardly improve overall map accuracy.
We used semivariograms to assess the spatial structure in the residuals of the RF predictions, and found that (non-spatial) RF rather regression kriging was justified at both sites.

@@

For both sites, we interpreted a model trained on a non-collinear subset of variables from the best performing predictor configuration -- because correlation between predictors degrades variable importance measures \citep{stroblConditionalVariableImportance2008, biauRandomForestGuided2016} and can produce misleading visualizations of predictor-outcome relationships \citep{biecekExplanatoryModelAnalysis2021, dwivediExplainableAIXAI2023}.
Specifically, we eliminated variables from the best performing predictor configuration to obtain a set with no pairwise Pearson correlation above 0.7.
Thus, highly-correlated sets of predictors are represented by a single variable for the purposes of model interpretation.

@@

Although the difficulty of predicting peat depth caused prediction intervals to be wide, these uncertainty estimates were well-calibrated.
At Skrimfjella the prediction interval coverage probability was 91 \%, and at Ørskogfjellet it was 84 \% (both compared to the target value of 90 \%).
Observations outside of the prediction intervals were evenly distributed across each of the study areas.

@@

For the purpose of model interpretation, the \emph{all predictors} configuration for Skrimfjella was reduced from 27 variables to 11 non-collinear variables.
Similarly, the \emph{terrain and DMK} configuration for Ørskogfjellet was reduced from 23 variables to 11 non-collinear variables.
At both sites, \emph{elevation} and \emph{MRVBF} were important predictors (Fig. \ref{fig:varImp}).
At Skrimfjella these two predictors were of similar importance, while at Ørskogfjellet \emph{elevation} was more important than \emph{MRVBF}.
\emph{DMK} was also important -- the shallow class in particular -- but only at Ørskogfjellet.
Some realizations of the hydrological predictors \emph{TWI} and \emph{DTW} showed considerable importance, while others showed little -- with no consistency between sites.
For example, \emph{TWI5m}, \emph{DTW40000}, and \emph{DTW2500} rounded out the top five most important variables at Skrimfjella, while \emph{TWI20m} and \emph{TWI50m} were least important, other than \emph{DMK}.
At both sites, the most important realizations of hydrological predictors were more important than the simple terrain indices \emph{slope}, \emph{TRI}, \emph{TPI}, and \emph{roughness}.
The radiometric variable \emph{radU} showed moderate importance at Skrimfjella.

@@

DMK peat depth classes were a worse predictor of peat depth than our models even though we calibrated them with the same data.
That is: we gave DMK and the DSM equal footing for a fair comparison.
If we had taken the DMK peat depth classes at face value and assumed depths according to their class definitions (\textless{} 1 m, \textgreater{} 1 m), they would have performed worse and the advantage of the DSM would be greater.
The advantage of the DSM was not large in absolute terms (9 cm and 21 cm improvements in mean absolute error), but it explained much more of the variation in depth (improvements in R\textsuperscript{2} of 0.16 and 0.19).
We attribute this result to the poor spatial and thematic resolution of DMK peat depth, which precludes a robust correlation with peat depths varying from 0-8 m at fine spatial scales.

>>>>

\begin{abstract}
Peatlands are Earth's most carbon-dense terrestrial ecosystems and their carbon density varies with the depth of the peat layer.
Accurate mapping of peat depth is crucial for carbon accounting and land management, yet existing maps lack the resolution and accuracy needed for these applications.
This study evaluates whether digital soil mapping using remotely sensed data can improve existing maps of peat depth in Norway.
Specifically, we assessed the predictive value of LiDAR-derived terrain variables and airborne radiometric data at two contrasting sites.
We measured peat depth at 372 and 1878 locations and trained random forest models to predict depth at 10 m resolution.
The best models achieved mean absolute errors of 60 and 56 cm, explaining one-third of the variation in peat depth.
Terrain variables were better predictors than radiometric variables, with elevation and valley bottom flatness showing the strongest relationships to depth.
Radiometric data showed inconsistent value, improving predictions at one site while degrading performance at the other.
Our models improved upon the national peat depth map, but weak relationships with remotely sensed variables make peat depth hard to predict.
These findings suggest that while digital soil mapping can enhance existing peat depth maps, reliable carbon stock assessments still require strategic field sampling.
Furthermore, the inability of models to identify peat presence outside known peatlands highlights the need for integrated mapping of peat occurrence and depth.
\end{abstract}

@@

Peat soils are a terrestrial carbon stock of global importance.
They store about 30 \% of global soil carbon despite covering only 2--3 \% of Earth's land \citep{xuPEATMAPRefiningEstimates2018, friedlingsteinGlobalCarbonBudget2020, unepGlobalPeatlandsAssessment2022}.
In other words, they are extremely carbon dense -- denser than any other ecosystem per square meter \citep{temminkRecoveringWetlandBiogeomorphic2022}.
This makes peatlands crucial to climate change mitigation.
Intact peatlands sequester carbon and produce a negative temperature forcing, overall \citep{joostenRolePeatlandsClimate2016}.
When disturbed, often by conversion to another land use, they can produce large greenhouse gas emissions \citep{maGloballyRobustRelationship2022}.

@@

The most practical and widespread \emph{scorpan} factors for peat depth are relief and spatial position.
Spatial position is unique because it is always known (with varying accuracy).
However, the short range of spatial autocorrelation in peat depth limits its mapping value \citep{henglGenericFrameworkSpatial2004}.
Relief, or topography, is widely and accurately mapped in digital terrain models (DTMs).
For example, most of mainland Norway has at least one elevation measurement per square meter from airborne Light Detection and Ranging (LiDAR) surveys.
Moreover, mechanisms of peat formation are linked to topography.
For example, a steep slope is unlikely to have a high water table, thereby limiting accumulation.

@@

The little we know about the depths of Norwegian peatlands comes from surveys meant to identify arable land.
After scattered surveys in the early 20th century, a comprehensive round of surveying was completed 1964--2001 as part of a wider land cover mapping in Norway \citep{bjordalMarkslagsklassifikasjonOkonomiskKartverk2007}.
Because of its agricultural and silvicultural focus, it covered only productive areas below the tree line, and peat depth was measured only in places judged to be potentially arable or afforestable \citep{ahlstromAR5Klassifikasjonssystem2019}.
Field surveyors carried a 1 m probe, so they measured peat depth as shallow (\textless{} 1 m), or deep (\textgreater{} 1 m).
These classes were assigned to whole mire polygons, so spatial resolution is on the order of hectares.

@@

At both study sites, our measurements of peat depth were made for the purpose of training a Random Forest (RF) model of peat depth, and we designed our sampling with this in mind \citep{brusSamplingDigitalSoil2019}.
Broadly, we aimed for a sample that was representative of the predictor space defined by the most important predictors of peat depth \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
A sample that preserves the properties of the multivariate distribution of predictor and outcome variables is most likely to maintain any complex, non-linear relationships that exist in the population while avoiding spurious ones \citep{brusSamplingDigitalSoil2019}.
We chose for our sampling and modelling a spatial resolution of \unit{10\,m}.
We considered this a reasonable compromise between DTM resolution (\unit{1\,m}) and small mires on the one hand, and airborne radiometric resolution (\unit{50\,m}) on the other.
The point measurements of peat depth described below were ultimately aggregated to \unit{10\,m} resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.

@@

Field work at Skrimfjella was conducted in August 2020.
We navigated to the centers of the raster cells in the depth and occurrence samples with a handheld global navigation satellite system (GNSS) receiver, checking that positional error was below \unit{3\,m}.
For each depth sample location, we measured peat depth three times (at the vertices of a triangle with \unit{2\,m} sides) to get a more representative value for the \unit{10\,m} raster cell, and to dampen the effect of outlying measurements \citep{parryEvaluatingApproachesEstimating2014}.
We used a metal probe pushed downward until resistance indicated the base of the peat column.
Probe locations were adjusted up to \unit{20\,cm} if the base of the peat column seemed to be blocked by an obvious artifact.
A single probe measurement was right-censored because the peat column was deeper than the probe.
For each occurrence sample location, we recorded the presence or absence of peatland -- primarily by digging and examining the top \unit{20\,cm} of soil (where this was possible).
We judged whether the soil was a peat soil based on its density, texture, and color.
Occasionally, when the soil itself was difficult to judge, we made our determination also based on the presence or absence of mire vegetation.
Although peat soil is strictly defined by organic content (which we did not analyze), we believe our protocol produced reasonable determinations of presence or absence that would generally satisfy most of the varying definitions of peatland \citep{minasnyMappingMonitoringPeatland2023}.

Besides the depth and occurrence measurements described above, we also measured peat depth in three subjectively chosen, individual mires, using ground-penetrating radar (GPR).
We used the Malå ProEx GPR system (Guideline Geo AB, Sweden) with its \unit{500\,MHz} shielded antenna mounted in a plastic sledge, and its control unit connected to a GNSS receiver.
At each of the three mires we recorded GPR traces along walking transects that covered the extent of the mire, mostly in traversing, zigzag patterns with between \unit{5\,m} and \unit{20\,m} spacing at their vertices.
Along the GPR transects we also probed peat depth at marked trace locations, to be able to calibrate the GPR wave speed velocity.
We processed the GPR data with Reflex2DQuick software (v.3.0, Sandmeier Scientific Software, Germany), applying a time-zero correction, a dewow filter, and a gain filter based on observed energy decay.
Then we picked strong reflectors in the radargrams that we interpreted as the base of the peat column.
We used picks at marked trace locations to calibrate wave speed velocity; we pooled calibration points across the three mires and fitted a linear regression of depth on one-way travel time with the intercept fixed at zero.
In total we had 46 calibration points along \unit{3.5\,km} of GPR transects.
Finally, we used the calculated wave velocity (\unit{0.0387\,m\,ns^{-1}}, \(R^2 = 0.874\)) to convert the travel times of all picks to calibrated peat depths.

@@

At Ørskogfjellet we also measured peat depth in a sample of \unit{10\,m} raster cells, selected from the part of the study area classified in the AR5 dataset as mire.
Before selecting locations, we determined a minimal sample size that would adequately capture the terrain and radiometric properties of the entire mire area.
Specifically, we aimed to identify the size at which adding locations produced diminishing decreases in divergence between sample and population distributions -- i.e.~the elbow point in a curve of similarity between sample and population \citep{maloneMethodsImproveUtility2019}.
This approach has been found to identify sample sizes that correspond with diminishing returns in predictive model performance on external evaluation data \citep{sauretteDivergenceMetricsDetermining2023}.
We defined a sequence of sample sizes (50--500) and for ten replicate samples at each size \citep[drawn by conditioned latin hypercube sampling,][]{minasnyConditionedLatinHypercube2006, roudierClhsPackageConditioned2011}, we calculated the mean Kullback--Leibler divergence between sample and population distributions \citep{maloneMethodsImproveUtility2019, sauretteDivergenceMetricsDetermining2023}.
The predictors in the divergence calculation were terrain slope and four radiometrics: potassium, thorium, uranium, and total count.
Next, we fitted a asymptotic regression of mean divergence on sample size, and identified the sample size at which the curve reached 95 \% of the fitted asymptote.
Through this procedure we found that we could adequately capture the population distribution with a sample of 160 locations.

To choose 160 locations, we performed feature space coverage sampling.
This approach has been found to produce higher accuracy in RFs than conditioned latin hypercube sampling \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
Feature space coverage sampling aims to disperse samples as uniformly as possible in multidimensional predictor space, and is implemented by choosing locations that are closest to cluster centers in a k-means clustering of the standardized predictor space \citep{brusSamplingDigitalSoil2019}.
Feature space coverage sampling works best when all dimensions are important predictors of the outcome \citep{wadouxSamplingDesignOptimization2019}, and we used the same five predictors that we used to choose sample size: terrain slope and four radiometrics.
The radiometric predictors were downscaled to \unit{10\,m} resolution with cubic B-spline resampling in QGIS \citep[v.3,][]{QGISsoftware}.
We adjusted the feature space coverage sampling to ensure that locations were accessible within time constraints, and assessed how this changed our sample from an ideal feature space coverage sample.
Adjusting for accessibility is justified because the smaller sample size that would result if accessibility were ignored can degrade model accuracy as much or more as deviations from ideal sampling designs \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
To adjust, we first restricted the sampling population to mire areas that were within an arbitrary cost distance of publicly accessible roads.
Cost distance was calculated using GRASS's \emph{r.walk} function, with friction costs defined by AR5 land classes \citep{GRASSv8-2}.
After creating a feature space coverage sample with this restriction, we also inspected a map of the sample and substituted 16 inaccessible locations with accessible locations from the same or a nearby cluster.
Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 \% (with respect to the ideal sample), but distance in our sample was still only 46 \% of the mean distance to cluster centers -- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.

@@

We used the \emph{terra} R package to calculate from the DTMs: slope, topographic position index (difference from mean of eight neighbors), terrain ruggedness index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
These were derived at two scales to produce eight different predictors; we either calculated from \unit{1\,m} DTM resolution and then aggregated, or aggregated to \unit{10\,m} DTM resolution and then calculated the indices.
This kind of multiscale feature engineering of land surface parameters has been found to improve machine learning predictions of soil properties \citep{millerImpactMultiscalePredictor2015, dornikOptimalScalingPredictors2022, newmanAssessingSpatiallyHeterogeneous2023}.
We know that peat depth tends to vary at fine scales in Norway, which is why we chose \unit{1\,m} and \unit{10\,m} resolutions \citep{maxwellLandsurfaceParametersSpatial2022}.
We also calculated the MRVBF index, which indicates the degree of valley bottom flatness at a given location via a multiscale algorithm \citep{gallantMultiresolutionIndexValley2003}.
We calculated this index in SAGA GIS \citep[v.9.3.2,][]{conradSystemAutomatedGeoscientific2015} with default parameters.

Next, we calculated the Topographic Wetness Index \citep{quinnPredictionHillslopeFlow1991}.
This index is notoriously scale-dependent and often matches real hydrological conditions best when calculated from moderate to coarse resolution DTMs \citep{agrenEvaluatingDigitalTerrain2014, riihimakiTopographicWetnessIndex2021}, so we calculated it from \unit{5\,m}, \unit{10\,m}, \unit{20\,m}, and \unit{50\,m} DTM resolution.
The calculations were performed with Whitebox software \citep{lindsayWhiteboxGATCase2016}, accessed through the \emph{whitebox} R package \citep[v2.4,][]{wuWhiteboxWhiteboxToolsFrontend2022}.
We filled depressions in the DTM with the algorithm in Wang \& Liu \citeyearpar{wangEfficientMethodIdentifying2006}, and used the deterministic infinity flow accumulation algorithm \citep{tarbotonNewMethodDetermination1997}.

The last terrain-based predictor we included was the depth-to-water (DTW) index \citep{murphyMappingWetlandsComparison2007}.
This index approximates a location's vertical height above the surface water feature that it is likely to drain towards.
It is calculated as the minimum cumulative slope (scaled by cell size) to a surface water feature \citep[eq. 5 in][]{murphyTopographicModellingSoil2009}.
We calculated unitless slope from the \unit{1\,m} DTM using the Whitebox software.
Also using Whitebox, we defined surface water features from the DTM by filling depressions and then calculating flow accumulation to define catchment areas for each cell \citep{schonauerSpatiotemporalPredictionSoil2021, schonauerRcodeCalculatingDepthwater2021}.
This catchments area layer was then thresholded at seven different levels (\emph{flow initiation areas} from \unit{0.25\,ha} to \unit{16\,ha}) to estimate surface water features under moisture scenarios varying from wet to dry \citep{murphyModellingMappingTopographic2011, agrenEvaluatingDigitalTerrain2014, schonauerSpatiotemporalPredictionSoil2021}.
In addition, all surface water features mapped in the AR5 dataset were also transferred to the raster layer.
For each of the seven surface water layers, we derived DTW using the \emph{Distance Accumulation} tool in ArcGIS Pro (v.3.1, ESRI, USA), which has an efficient algorithm to find the cumulative distance over a cost surface to the least-cost source.

@@

We used RF to predict peat depth at both sites.
RF is a tree-based ensemble machine learning algorithm that builds many decision trees on bootstrapped samples of the training data, randomly subsets predictors in the trees, and averages the predictions of the trees \citep{breimanRandomForests2001}.
We chose RF because it can handle complex interactions between predictors, is robust to overfitting, and generally shows higher performance in DSM applications than other algorithms \citep{beguinPredictingSoilProperties2017, nussbaumEvaluationDigitalSoil2018, lamichhaneDigitalSoilMapping2019}.
It is suited for use on relatively small training data sets and its predictions can be interrogated to learn about predictor importance \citep{khaledianSelectingAppropriateMachine2020}.
Evaluating variable importance in a maximally predictive model aligns with the aim of this study.

RF by itself is not a spatial model, and it will only predict spatial structure in the outcome to the degree that the structure is captured by predictors.
We considered using regression kriging -- a hybrid between non-spatial and spatial techniques that would be achieved by adding to the RF predictions a geostatistically interpolated surface of RF residuals \citep{henglGenericFrameworkSpatial2004}.
The spatial component in regression kriging often improves map accuracy compared to a non-spatial model \citep{beguinPredictingSoilProperties2017, lamichhaneDigitalSoilMapping2019, mollaMachineLearningGeostatistical2023}, but it can do so only if the spatial autocorrelation range in the non-spatial residuals is large compared to distances between samples and prediction locations \citep{henglGenericFrameworkSpatial2004, szaboMappingSoilHydraulic2019, takoutsingComparingPredictionPerformance2022}.
If the outcome varies at fine scales and the samples are clustered in small parts of the study area, a spatial component will hardly improve overall map accuracy.
We used semivariograms to assess the spatial structure in the residuals of the RF predictions, and found that (non-spatial) RF rather regression kriging was justified at both sites.

@@

For both sites, we interpreted a model trained on a non-collinear subset of variables from the best performing predictor configuration -- because correlation between predictors degrades variable importance measures \citep{stroblConditionalVariableImportance2008, biauRandomForestGuided2016} and can produce misleading visualizations of predictor--outcome relationships \citep{biecekExplanatoryModelAnalysis2021, dwivediExplainableAIXAI2023}.
Specifically, we eliminated variables from the best performing predictor configuration to obtain a set with no pairwise Pearson correlation above 0.7.
Thus, highly correlated sets of predictors are represented by a single variable for the purposes of model interpretation.

@@

Although the difficulty of predicting peat depth caused prediction intervals to be wide, these uncertainty estimates were well calibrated.
At Skrimfjella the prediction interval coverage probability was 91 \%, and at Ørskogfjellet it was 84 \% (both compared to the target value of 90 \%).
Observations outside of the prediction intervals were evenly distributed across each of the study areas.

@@

For the purpose of model interpretation, the \emph{all predictors} configuration for Skrimfjella was reduced from 27 variables to 11 non-collinear variables.
Similarly, the \emph{terrain and DMK} configuration for Ørskogfjellet was reduced from 23 variables to 11 non-collinear variables.
At both sites, \emph{elevation} and \emph{MRVBF} were important predictors (Fig. \ref{fig:varImp}).
At Skrimfjella these two predictors were of similar importance, while at Ørskogfjellet \emph{elevation} was more important than \emph{MRVBF}.
\emph{DMK} was also important -- the shallow class in particular -- but only at Ørskogfjellet.
Some realizations of the hydrological predictors TWI and DTW showed considerable importance, while others showed little -- with no consistency between sites.
For example, \emph{TWI5m}, \emph{DTW40000}, and \emph{DTW2500} rounded out the top five most important variables at Skrimfjella, while \emph{TWI20m} and \emph{TWI50m} were least important, other than \emph{DMK}.
At both sites, the most important realizations of hydrological predictors were more important than the simple terrain indices slope, TRI, TPI, and roughness.
The radiometric variable \emph{radU} showed moderate importance at Skrimfjella.

@@

DMK peat depth classes were a worse predictor of peat depth than our models even though we calibrated them with the same data.
That is: we gave DMK and the DSM equal footing for a fair comparison.
If we had taken the DMK peat depth classes at face value and assumed depths according to their class definitions (\textless{} 1 m, \textgreater{} 1 m), they would have performed worse and the advantage of the DSM would be greater.
The advantage of the DSM was not large in absolute terms (9 cm and 21 cm improvements in mean absolute error), but it explained much more of the variation in depth (improvements in R\textsuperscript{2} of 0.16 and 0.19).
We attribute this result to the poor spatial and thematic resolution of DMK peat depth, which precludes a robust correlation with peat depths varying from 0--8 m at fine spatial scales.
